---
title: "Demographic Factors"
execute: 
  echo: false
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(brms)
library(tidybayes)
library(gtsummary)
library(wesanderson)
library(plotly)
```

```{r}
#| label: import_data

# Brought downloaded CSV file with data into project through terminal. Turned
# CSV file into data frame. Kept only the data that combined all age groups
# together to see the general trend in incidence rates over time.

ms_data_dem <- read_csv("DataFilterExport.csv", show_col_types = FALSE)
ms_data_clean_time <- ms_data_dem |>
  filter(Age == "ALL")
ms_data_clean <- ms_data_dem |>
  filter(Year == "2022") 
```

## MS Incidence Rate Over Time By Sex

```{r}
#| label: create_line_graph

# Created line graph to show trend of MS Incidence Rates in Alberta over time.
# Tried bar graph but decided on line graph so trend was more apparent. Created
# lines for each of the Sex categories to see the effect of Sex on incidence
# rate change over time. Used Wes Anderson Palettes R package for colors in line
# graph. Opted not to add title since it was the same as the page and subtitle
# since key takeaways were listed in paragraph below. Used minimal theme for
# clean look.

ms_line <-
  ms_data_clean_time |>
  ggplot(aes(x = Year, y = `Incidence Rate`, color = Sex)) +
  geom_line(linewidth = 0.75) +
  scale_color_manual(values = wes_palette("Cavalcanti1")[c(4, 1, 5)]) +
  labs(x = "Year", 
       y = "Incidence Rate (per 100 000)", 
       color = "Sex") +
  theme_minimal()

# Used plotly package to make graph interactive. Added caption to provide source
# of data, and adjusted sizing and placement of caption. Displayed line graph.

interactive_line <- 
  ggplotly(ms_line, tooltip = c("Year", "Incidence Rate")) |>
    layout( annotations = list(
      text = "Data from Alberta IHDA",
      x = 1, y = -0.25,
      xref = "paper", yref = "paper",
      showarrow = FALSE,
      font = list(size = 12)), 
      margin = list(b = 100))

interactive_line
```

Using MS Incidence Rate data, which is the number of incident (new) cases divided by the total population at risk per 100 000 people, from the government of Alberta’s Interactive Health Data Application, it appears that women are diagnosed with MS at a higher rate than men. Using data from the 2006-2010 study, I also explored the relationship between sex and MS diagnosis. 


```{r}
raw_df <- read_csv("ms-conversion-rate.csv", show_col_types = FALSE)

ms_data <- raw_df[-c(274:288) ,] |>
  mutate(Age = `Age (y)`) |>
  mutate(Schooling = `Schooling (y)`) |>
  mutate(Gender = recode(Gender,
                         '1' = "male",
                         '2' = "female")) |>
  mutate(Breastfeeding = recode(Breastfeeding,
                                '1' = "yes",
                                '2' = "no",
                                '3' = "unknown")) |>
  mutate(Varicella = recode(Varicella,
                            '1' = "positive",
                            '2' = "negative",
                            '3' = "unknown")) |>
  mutate(Initial_Symptoms = recode(`initial symptom`,
                                  '1' = "visual",
                                  '2' = "sensory",
                                  '3' = "motor",
                                  '4' = "other",
                                  '5' = "visual  and sensory",
                                  '6' = "visual  and motor",
                                  '7' = "visual  and other",
                                  '8' = "sensory  and motor",
                                  '9' = "sensory  and other",
                                  '10' = "motor  and  other",
                                  '11' = "visual,  sensory and motor",
                                  '12' = "visual, sensory  and other",
                                  '13' = "visual, motor and  other",
                                  '14' = "sensory,  motor and   other",
                                  '15' = "visual,sensory,motor and other")) |>
  mutate(Mono_Poly = recode(`Mono or polysymptomatic`,
                            '1' = "monosymptomatic",
                            '2' = "polysymptomatic",
                            '3' = "unknown")) |>
  mutate(Oligoclonal_Bands = recode(`Oligoclonal bands`,
                                   '0' = "negative",
                                   '1' = "positive",
                                   '2' = "unknown")) |>
  mutate(LLSSEP = recode(LLSSEP,
                        '0' = "negative",
                        '1' = "positive")) |>
  mutate(ULSSEP = recode(ULSSEP,
                        '0' = "negative",
                        '1' = "positive")) |>
  mutate(VEP = recode(VEP,
                      '0' = "negative",
                      '1' = "positive")) |>
  mutate(BAEP = recode(BAEP,
                       '0' = "negative",
                       '1' = "positive")) |>
  mutate(Periventricular_MRI = recode(`Periventricular MRI`,
                                      '0' = "negative",
                                      '1' = "positive")) |>
  mutate(Cortical_MRI = recode(`Cortical MRI`,
                                '0' = "negative",
                                '1' = "positive")) |>
  mutate(Infratentorial_MRI = recode(`Infratentorial MRI`,
                                      '0' = "negative",
                                      '1' = "positive")) |>
  mutate(Spinal_Cord_MRI = recode(`Spinal cord MRI`,
                                  '0' = "negative",
                                  '1' = "positive")) |> 
  mutate(group = recode(group,
                        '1' = "CDMS",
                        '2' = "Non-CDMS")) |>
  select(Patient, 
         Gender, 
         Age, 
         Schooling, 
         Breastfeeding, 
         Varicella,
         Initial_Symptoms,
         Mono_Poly,
         Oligoclonal_Bands,
         LLSSEP,
         ULSSEP,
         VEP,
         BAEP,
         Periventricular_MRI,
         Cortical_MRI,
         Infratentorial_MRI,
         Spinal_Cord_MRI,
         group) |> 
  drop_na()
```

## Statistical Model

```{r}
#| cache: true
fit_dem <- brm(formula = group ~ Gender,
              data = ms_data,
              family = bernoulli(),
              silent = 2,
              refresh = 0,
              seed = 9)
```


```{r}
#| message: false
ndata <- tibble(Gender = c("male", "female"))

plot <- fit_dem |>
  add_epred_draws(newdata = ndata) |>
  ungroup() |>
  ggplot(aes(x = .epred, fill = Gender)) + 
  geom_density(aes(y = after_stat(count/sum(count)))) +
  labs(title = "Posterior Distribution for Probability of Developing MS Based on Sex",
       x = "Expected Probability of Developing MS",
       y = "Probability",
       fill = "Sex") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = wes_palette("AsteroidCity1")) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_classic()

interactive_plot <- ggplotly(plot)

interactive_plot
```

```{r}
tbl_regression(fit_dem)
```

Men seem to have about 63% lower odds of being diagnosed with MS than women, although that estimate could be as high as 77% or as low as 43%. I modeled sex, with only values of "male" and "female", as a logistic function of whether or not a patient was diagnosed with MS. Patients who were male were less likely to be diagnosed with MS than those who were female.


## Additional Graphs 

```{r}
#| label: create_bar_graph

# Graphed Age vs Incidence Rate for each Sex. Tried to create faceted bar graph
# using facet_wrap() but decided it would be more valuable to see data for each
# category next to each other. Added error bars using Standard Error from data
# set. Used Wes Anderson Palettes R Package for the colors in the bar graph.
# Opted not to add subtitle since key takeaways were listed in paragraph below.
# Used minimal theme for clean look. Adjusted the labels on the x-axis so they
# were easy to read.

ms_bar <- 
  ms_data_clean |> 
  drop_na() |>
  ggplot(aes(x = Age, y = `Incidence Rate`, fill = Sex)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  geom_errorbar(aes(ymin = `Incidence Rate` - `Standard Error`,
                    ymax = `Incidence Rate` + `Standard Error`),
                position = position_dodge(width = 0.7), 
                width = 0.25,  
                alpha = 0.75) +
  scale_fill_manual(values = wes_palette("AsteroidCity1")) +
  labs(x = "Age Group", y = "Incidence Rate (per 100 000)", 
       title = "Age-Sex Specific Incidence Rates in Alberta, 2022",
       fill = "Sex") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  


# Used plotly package to make graph interactive. Added caption to provide source
# of data, and adjusted sizing and placement of caption. Displayed bar graph.

interactive_bar <- 
  ggplotly(ms_bar, tooltip = c("Age", "Incidence Rate")) |>
    layout(
      annotations = list(
        text = "Data from Alberta IHDA",
        x = 1, y = -0.25,
        xref = "paper", yref = "paper",
        showarrow = FALSE,
        font = list(size = 12)
      ), 
      margin = list(b = 100),
      title = list(
        text = "Age-Sex Specific Incidence Rates in Alberta, 2022<br><sub>Women tend to be diagnosed more often than men, indicating that sex might be a risk factor for developing MS</sub>",
        yanchor = 'top'
      )
    )

interactive_bar
```

```{r}
#| label: create_bar

# Graphed Age vs Incidence Rate for each Sex for all years. Tried to create
# faceted bar graph using facet_wrap() but decided it would be more valuable to
# see data for each category next to each other. Used Wes Anderson Palettes R
# Package for the colors in the bar graph. Opted not to add title since it was
# the same as the page and subtitle since key takeaways were listed in paragraph
# below. Used minimal theme for clean look. Adjusted the labels on the x-axis so
# they were easy to read.

ms_bar_all <- 
  ms_data_dem |>
  drop_na() |>
  ggplot(aes(x = Age, y = `Incidence Rate`, fill = Sex)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  scale_fill_manual(values = wes_palette("Darjeeling1")[c(2, 3, 1)]) +
  labs(x = "Age Group", y = "Incidence Rate (per 100 000)", 
       fill = "Sex", title = "Age-Sex Specific Incidence Rates, 2004-2022") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  


# Used plotly package to make graph interactive. Added caption to provide source
# of data, and adjusted sizing and placement of caption. Displayed bar graph.

interactive_bar_all <- 
  ggplotly(ms_bar_all, tooltip = c("Age", "Incidence Rate")) |>
    layout(
      annotations = list(
        text = "Data from Alberta IHDA",
        x = 1, y = -0.25,
        xref = "paper", yref = "paper",
        showarrow = FALSE,
        font = list(size = 12)
      ), 
      margin = list(b = 100),
      title = list(
        text = "Age-Sex Specific Incidence Rates, 2004-2022<br><sub>The trends across all the years of data available seems to be consistent with the data in 2022</sub>",
        yanchor = 'top'
      )
    )

interactive_bar_all
```
