---
title: "Conversion from CIS to MS"
author: "Ashwika Katiyar"
execute: 
  echo: false
---

```{r}
#| label: setup
#| message: false 

library(tidyverse)
library(brms)
library(tidybayes)
library(gtsummary)
library(wesanderson)
```

```{r}
#| label: download_format_data

# Brought downloaded CSV file with data into project through
# terminal. Turned CSV file into data frame.

raw_df <- read_csv("ms-conversion-rate.csv", show_col_types = FALSE)

# Recoded columns from numbers into descriptive labels using key at the bottom of
# the data CSV.

ms_data <- raw_df[-c(274:288) ,] |>
  mutate(Age = `Age (y)`) |>
  mutate(Schooling = `Schooling (y)`) |>
  mutate(Gender = recode(Gender,
                         '1' = "male",
                         '2' = "female")) |>
  mutate(Breastfeeding = recode(Breastfeeding,
                                '1' = "yes",
                                '2' = "no",
                                '3' = "unknown")) |>
  mutate(Varicella = recode(Varicella,
                            '1' = "positive",
                            '2' = "negative",
                            '3' = "unknown")) |>
  mutate(Initial_Symptoms = recode(`initial symptom`,
                                  '1' = "visual",
                                  '2' = "sensory",
                                  '3' = "motor",
                                  '4' = "other",
                                  '5' = "visual  and sensory",
                                  '6' = "visual  and motor",
                                  '7' = "visual  and other",
                                  '8' = "sensory  and motor",
                                  '9' = "sensory  and other",
                                  '10' = "motor  and  other",
                                  '11' = "visual,  sensory and motor",
                                  '12' = "visual, sensory  and other",
                                  '13' = "visual, motor and  other",
                                  '14' = "sensory,  motor and   other",
                                  '15' = "visual,sensory,motor and other")) |>
  mutate(Mono_Poly = recode(`Mono or polysymptomatic`,
                            '1' = "monosymptomatic",
                            '2' = "polysymptomatic",
                            '3' = "unknown")) |>
  mutate(Oligoclonal_Bands = recode(`Oligoclonal bands`,
                                   '0' = "positive",
                                   '1' = "negative",
                                   '2' = "unknown")) |>
  mutate(LLSSEP = recode(LLSSEP,
                        '0' = "negative",
                        '1' = "positive")) |>
  mutate(ULSSEP = recode(ULSSEP,
                        '0' = "negative",
                        '1' = "positive")) |>
  mutate(VEP = recode(VEP,
                      '0' = "negative",
                      '1' = "positive")) |>
  mutate(BAEP = recode(BAEP,
                       '0' = "negative",
                       '1' = "positive")) |>
  mutate(Periventricular_MRI = recode(`Periventricular MRI`,
                                      '0' = "negative",
                                      '1' = "positive")) |>
  mutate(Cortical_MRI = recode(`Cortical MRI`,
                                '0' = "negative",
                                '1' = "positive")) |>
  mutate(Infratentorial_MRI = recode(`Infratentorial MRI`,
                                      '0' = "negative",
                                      '1' = "positive")) |>
  mutate(Spinal_Cord_MRI = recode(`Spinal cord MRI`,
                                  '0' = "negative",
                                  '1' = "positive")) |> 
  mutate(group = recode(group,
                        '1' = "CDMS",
                        '2' = "Non-CDMS")) |>
  select(Patient, 
         Gender, 
         Age, 
         Schooling, 
         Breastfeeding, 
         Varicella,
         Initial_Symptoms,
         Mono_Poly,
         Oligoclonal_Bands,
         LLSSEP,
         ULSSEP,
         VEP,
         BAEP,
         Periventricular_MRI,
         Cortical_MRI,
         Infratentorial_MRI,
         Spinal_Cord_MRI,
         group) |> 
  drop_na()


# Removed rows with "unknown" for Oligoclonal Bands for clarity

x <- ms_data |>
  filter(Oligoclonal_Bands %in% c("positive", "negative")) 

set.seed(123)
x <- x[sample(nrow(x), size = 1000, replace = TRUE), ]
```

```{r}
#| label: model

# Created brm model to compare group and Oligoclonal Bands. Used bernoulli
# family because group and Oligoclonal_Bands are binary variables. Used silent
# and refresh to get rid of extra output. Used seed to get replicable results.

fit_ob <- readRDS("fit_ob.rds")
```


```{r}
#| label: posterior_plot

# Created ndata object with all possible values for the right side variable

ndata <- tibble(expand_grid(Oligoclonal_Bands = c("positive", "negative")))

# Graphed posterior distribution of model. Initially tried making a density plot
# but decided histogram better displayed the data. Used Wes Anderson Palettes
# R Package for the colors in the bar graph. Opted not to add subtitle since key
# takeaways were listed in paragraph below. Used minimal theme for clean look.

fit_ob |>
  add_epred_draws(newdata = ndata) |>
  ungroup() |>
  ggplot(aes(x = .epred, fill = Oligoclonal_Bands)) + 
  geom_histogram(aes(y = after_stat(count/sum(count))), bins = 50) +
  labs(title = "Posterior Distribution for Probability of Developing MS Based on OCBs",
       x = "Expected Probability of Developing MS",
       y = "Probability",
       fill = "Presence of OCBs") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = wes_palette("FrenchDispatch"),
                    labels = c("negative" = "Positive", "positive" = "Negative")) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_classic()

```

**Introduction:**
Multiple Sclerosis (MS) is a devastating, demyelinating disease of the central nervous system that affects people of all ages and genders, and Clinically Isolated Syndrome (CIS) is a patient's first episode of neurological symptoms, which may then develop into MS. 

**Data:**
Using data from a study of the medical histories of Mexican mestizo patients diagnosed with CIS at NINN in Mexico City from 2006 to 2010, I sought to understand the connection between various symptoms and factors and the patient's later diagnosis of MS. 

After exploring the data, I focused on the presence of Oligoclonal Bands (OCBs), which are bands of immunoglobulins seen in blood serum or cerebrospinal fluid, and isolated the data with only the responses "positive" and "negative". Positive indicates detected presence while negative indicates OCBs were not detected. 

**Conclusion:**
Patients without a presence of OCBs have about 84.5% lower odds to be diagnosed with MS, although that could be as low as 80% or as high as 89%. 

**Model:**
Although the data included multiple variables such as age, gender, years of schooling, and past Varicella infections, I chose to model Oligoclonal Bands, with only values of "positive" and "negative", as a logistic function of whether or not a patient was diagnosed with MS. Patients with a positive presence were less likely to be diagnosed with MS than those with a negative presence.  


